---
title: "Causal win statistics for hierarchical composite endpoints"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal win statistics for hierarchical composite endpoints}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
set.seed(1)
```

## What `mrwin` estimates

`mrwin` targets a **dose-standardised causal win ratio (DS-CWR)** for a
hierarchical composite endpoint in a Mendelian-randomization (MR) design. The
outcome is several priority-ordered events (e.g. death > hospitalisation >
biomarker decline); a polygenic score (PRS) built from GWAS weights is the
instrument for an exposure. The estimator:

1. ranks subjects by the PRS and bins them into strata;
2. for each adjacent stratum pair, forms the hierarchical win/loss log-odds
   `log theta` and the exposure shift `Delta_X`, and takes their ratio — the
   **instrument-standardised gradient** `ISG = log theta / Delta_X`;
3. pools the adjacent ISGs by generalised least squares to get `delta_GLS`;
   `DS-CWR = exp(delta_GLS)`.

`DS-CWR > 1` means the genetically-predicted exposure shifts subjects toward a
**better** prioritised outcome profile, `< 1` toward a worse one.

## A worked example on simulated data

The package ships a simulation engine. We draw a cohort with a genuine
protective exposure effect (`alpha_x < 0`) and confounding.

```{r simulate}
library(mrwin)

cfg <- mrwin_config(n_outcome = 3000L, m_snps = 50L, seed = 1L)
dat <- mrwin_simulate(cfg, seed = 1L)

endpoint <- mrwin_endpoint(
  time = dat$time, status = dat$status,
  priority = c("death", "hospitalisation", "biomarker")
)
gwas <- mrwin_gwas(beta = dat$true_betas, se = rep(0.01, length(dat$true_betas)))
```

Fit the model. The default backend is dense and the default inference is the
multiplier bootstrap.

```{r fit}
fit <- mrwin(
  endpoint = endpoint, genotype = dat$G, exposure = dat$X, gwas = gwas,
  controls = mrwin_controls(n_strata = 5L, bootstrap = 100L, seed = 1L,
                            run_sdpd = FALSE)
)
summary(fit)
```

The reported interval is the **Fieller** interval — the calibrated choice for
this ratio estimand (see below). A tidy one-row summary is also available:

```{r tidy}
tidy(fit)
```

```{r plot, fig.alt = "Adjacent-stratum instrument-standardised gradients"}
plot(fit, type = "isg")
```

## Biobank-scale backend

The hierarchical win/loss kernel is the bottleneck. `backend = "fast"` uses a
subquadratic, compiled (Rcpp) kernel that is **bit-for-bit identical** to the
dense backend but scales to biobank `N`. The point estimate is unchanged:

```{r fast}
fit_fast <- mrwin(
  endpoint = endpoint, genotype = dat$G, exposure = dat$X, gwas = gwas,
  controls = mrwin_controls(n_strata = 5L, bootstrap = 100L, seed = 1L,
                            backend = "fast", run_sdpd = FALSE)
)
c(dense = unname(fit$point$dscwr), fast = unname(fit_fast$point$dscwr))
```

## Analytic variance (a fast alternative to the bootstrap)

`inference = "analytic"` replaces the bootstrap loop for the sampling component
with a closed-form influence-function covariance (with an exact Monte-Carlo term
for GWAS-weight uncertainty when `se > 0`).

```{r analytic}
fit_analytic <- mrwin(
  endpoint = endpoint, genotype = dat$G, exposure = dat$X, gwas = gwas,
  controls = mrwin_controls(n_strata = 5L, seed = 1L,
                            inference = "analytic", run_sdpd = FALSE)
)
fit_analytic$inference$ci95_dscwr_fieller
```

## Why Fieller, and the weak-instrument case

The ISG is a ratio. External calibration showed that a delta-method interval for
it is structurally too wide (it never reaches nominal type-I error), whereas the
**Fieller** construction on the same covariance is correctly calibrated. `mrwin`
therefore reports the Fieller interval as primary and keeps the delta-method
interval only as a labelled reference. Under a weak instrument the Fieller
interval is honestly reported as unbounded rather than as a falsely tight set.

## A caveat: doubly-ranked stratification

Doubly-ranked stratification is popular in non-linear MR, but it **balances the
instrument across strata**, which makes the *between*-stratum DS-CWR contrast
confounder-driven (validated type-I error ~1.0). `mrwin()` keeps it runnable for
research but warns; `prs_rank` (the default) is the only validated scheme.

```{r doubly-ranked}
fit_dr <- tryCatch(
  withCallingHandlers(
    mrwin(endpoint = endpoint, genotype = dat$G, exposure = dat$X, gwas = gwas,
          controls = mrwin_controls(n_strata = 5L, bootstrap = 50L, seed = 1L,
                                    stratification = "doubly_ranked",
                                    run_sdpd = FALSE)),
    warning = function(w) {
      message("warning: ", conditionMessage(w))
      invokeRestart("muffleWarning")
    }
  )
)
```

See `vignette` sources and `inst/spec/validation-findings.md` for the full
calibration evidence behind these choices.
